Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-02-28 paperDistance/paper.tex

% # To-do list
% - zeroth draft the introduction
% - zeroth draft the method section
% - zeroth draft the assumptions list
% - reformat single-column

@@

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
We obtained these by building a data-driven model of the
color--magnitude diagram (\cmd) of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian interences.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters [HOGG: Is this too general?] are much more clustered than likelihood-based distance
estimates.

@@

What would be different if we wanted a model not of the \tgas\ data,
but a model appropriate for the full billion stars, or for a
volume-limited sample of stars?
How could we convert our existing results?
How would we do it properly in general?

>>>>

% # To-do list
% - We need a standard name for the \tgas--\tmass\ Intersection.
% - zeroth draft the introduction
% - zeroth draft the method section
% - zeroth draft the assumptions list
% - reformat single-column

@@

\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
That is, the geometric parallaxes can calibrate a set of photometric
models that are purely statistical---a model of the data rather than
a model of stars \foreign{per se}.
This opens up the possibility of completing the goals of the \gaia\ Mission
without building in unnecessary assumptions about the mechanical properties
of stars or the Galaxy.

We obtained the photometric parallaxes in this project
by building a data-driven model of the
color--magnitude diagram (\cmd) of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian interences.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters [HOGG: Is this too general?] are much more clustered than likelihood-based distance
estimates.

@@

In some sense, the biggest limitation of the data-driven \cmd\ model
created in this project is that we made no use of any kind of \tgas\ selection
function or completeness estimate or inverse-selection-volume corrections.
That is, the model is a model of the \tgas\ Catalog
(or really the \tgas--\tmass\ intersection), not of the stars in
the Milky Way, nor any volume-limited subsample of the Milky Way.
Importantly, because we used \emph{all} of the stars in the \tgas--\tmass\
intersection, the model \emph{is} representive of the full catalog,
even the distant parts, where all the parallaxes are measured at low
signal-to-noise.
This is in contrast to techniques that might build the model from only
the high signal-to-noise sources; such a model would be biased towards
stellar types and compositions found locally, lower-luminosity stars,
and stars which happened to get good parallaxes.
However, our use of the entire \tgas--\tmass\ intersection without
any selection function restricts the use of our \cmd\ model as a prior
to inferences within that same \tgas--\tmass\ intersection.
For example, it would be biasing to use this same prior to infer
distances for the full billion-star catalog released in \gaia\ DR1
alongside \tgas, or for stars fainter or brighter or bluer than those
included in the \tgas--\tmass\ intersection.

Note the dangers of using the stars from our output in hierarchical inferences!
